Using CopyKAT to identify tumor cells in Ewing sarcoma samples

### 2024-07-24 analyses/cell-type-ewings/exploratory_analysis/04-infercnv.Rmd
fix up some of the ewings files with diffs

png_files |> 
  # make sure png files exists
  purrr::walk(\(file)
    stopifnot("Missing PNG file" = file.exists(file))
  )

@@

# don't include the normal cells that were provided as reference in scaling 
cnv_df <- all_cnv_df |>
  # get total number of chromosomes with cnvs per cell and then scale that number 
  dplyr::mutate(
    has_cnv = ifelse(
      stringr::str_detect(subcluster, "reference"), 
      NA_real_, 
      rowSums(across(has_cnv_cols))
    ))
```

```{r}
# plot the total number of CNVs
ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = has_cnv)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  scale_color_viridis_c()
```
This shows that the total CNVs are highest in the group of cells that correspond to the same cells that are manually annotated as tumor cells, which is promising. 
Those cells in gray (bottom left) are cells used as a reference. 
If we were to use total CNVs we would need some sort of cut off for the number. 
Let's look at the distribution of total CNVs to see if we can identify a cut off.

@@

```{r}
# get mean value for has_cnv, excluding any NAs
mean_cnv <- mean(cnv_df$has_cnv, na.rm=TRUE)

# add cnv classification 
cnv_df <- cnv_df |> 
  dplyr::mutate(cnv_classification = dplyr::case_when(is.na(has_cnv) ~ "Reference",
                                                      has_cnv > mean_cnv ~ "Tumor",
                                                      has_cnv <= mean_cnv ~ "Normal"))

ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = cnv_classification)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() 
```

@@

# get mean proportion of cnvs per cell weighted by chromsome size (number of genes)
cnv_df <- cnv_df |> 
  dplyr::rowwise() |> 
  # calculate weighted mean, excluding normal cells 
  dplyr::mutate(
    mean_proportion = ifelse(
      !is.na(has_cnv), 
      weighted.mean(across(proportion_cols), chr_weights), 
      NA)) |> 
  dplyr::ungroup()
```

```{r}
ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = mean_proportion)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  scale_color_viridis_c()
```

@@

# add cnv classification 
cnv_df <- cnv_df |> 
  dplyr::mutate(cnv_mean_classification = dplyr::case_when(is.na(mean_proportion) ~ "Reference",
                                                           mean_proportion > all_mean_proportion ~ "Tumor",
                                                           mean_proportion <= all_mean_proportion ~ "Normal"))

ggplot(cnv_df, aes(x = UMAP1, y= UMAP2, color = cnv_mean_classification)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() 
```

@@

```{r, fig.height=10, fig.width=7}
# Set heatmap padding option
heatmap_padding <- 0.2
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal 
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(celltype_mat, celltype_method) {
      ComplexHeatmap::Heatmap(
        t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = celltype_method,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = celltype_method == "SingleR",
      )
    }) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

@@

```{r}
# prepare a data frame with one row per chromosome per cell for plotting
proportion_df <- cnv_df |> 
  # select only relevant columns for plotting
  dplyr::select(barcodes, UMAP1, UMAP2, cnv_mean_classification, all_of(proportion_cols)) |> 
  tidyr::pivot_longer(cols = all_of(proportion_cols),
    names_to = "chromosome",
    values_to = "proportion") |> 
  # remove extra string in front of chr 
  dplyr::mutate(
    chromosome = stringr::str_remove(chromosome, "proportion_scaled_cnv_chr"),
    # turn into numeric for easy ordering in plot 
    chromosome = as.numeric(chromosome)
  )

```

```{r, fig.height=10, fig.width=10}
ggplot(proportion_df, aes(x = proportion, color = cnv_mean_classification)) +
  geom_density(bounds = c(0,1), bw = 0.05) +
  theme_bw() +
  facet_wrap(vars(chromosome), ncol = 3)
```

@@

```{r}
ewings_chr_df <- proportion_df |> 
  dplyr::filter(chromosome %in% c(1,8,12,16))

ggplot(ewings_chr_df, aes(x = UMAP1, y=UMAP2 , color = proportion)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  facet_wrap(vars(chromosome)) +
  scale_color_viridis_c()

```

>>>>

png_files |>
  # make sure png files exists
  purrr::walk(\(file)
  stopifnot("Missing PNG file" = file.exists(file)))

@@

# don't include the normal cells that were provided as reference in scaling
cnv_df <- all_cnv_df |>
  # get total number of chromosomes with cnvs per cell and then scale that number
  dplyr::mutate(
    has_cnv = ifelse(
      stringr::str_detect(subcluster, "reference"),
      NA_real_,
      rowSums(across(has_cnv_cols))
    )
  )
```

```{r}
# plot the total number of CNVs
ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = has_cnv)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  scale_color_viridis_c()
```
This shows that the total CNVs are highest in the group of cells that correspond to the same cells that are manually annotated as tumor cells, which is promising. 
Those cells in gray (bottom left) are cells used as a reference. 
If we were to use total CNVs we would need some sort of cut off for the number. 
Let's look at the distribution of total CNVs to see if we can identify a cut off.

@@

```{r}
# get mean value for has_cnv, excluding any NAs
mean_cnv <- mean(cnv_df$has_cnv, na.rm = TRUE)

# add cnv classification
cnv_df <- cnv_df |>
  dplyr::mutate(cnv_classification = dplyr::case_when(
    is.na(has_cnv) ~ "Reference",
    has_cnv > mean_cnv ~ "Tumor",
    has_cnv <= mean_cnv ~ "Normal"
  ))

ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = cnv_classification)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw()
```

@@

# get mean proportion of cnvs per cell weighted by chromsome size (number of genes)
cnv_df <- cnv_df |>
  dplyr::rowwise() |>
  # calculate weighted mean, excluding normal cells
  dplyr::mutate(
    mean_proportion = ifelse(
      !is.na(has_cnv),
      weighted.mean(across(proportion_cols), chr_weights),
      NA
    )
  ) |>
  dplyr::ungroup()
```

```{r}
ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = mean_proportion)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  scale_color_viridis_c()
```

@@

# add cnv classification
cnv_df <- cnv_df |>
  dplyr::mutate(cnv_mean_classification = dplyr::case_when(
    is.na(mean_proportion) ~ "Reference",
    mean_proportion > all_mean_proportion ~ "Tumor",
    mean_proportion <= all_mean_proportion ~ "Normal"
  ))

ggplot(cnv_df, aes(x = UMAP1, y = UMAP2, color = cnv_mean_classification)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw()
```

@@

```{r, fig.height=10, fig.width=7}
# Set heatmap padding option
heatmap_padding <- 0.2
ComplexHeatmap::ht_opt(TITLE_PADDING = grid::unit(heatmap_padding, "in"))
# list of heatmaps looking at SingleR/ CellAssign vs tumor/normal
heatmap <- jaccard_matrices |>
  purrr::imap(
    \(celltype_mat, celltype_method) {
      ComplexHeatmap::Heatmap(
        t(celltype_mat), # transpose because matrix rows are in common & we want a vertical arrangement
        col = circlize::colorRamp2(c(0, 1), colors = c("white", "darkslateblue")),
        border = TRUE,
        ## Row parameters
        cluster_rows = TRUE,
        row_title = celltype_method,
        row_title_gp = grid::gpar(fontsize = 12),
        row_title_side = "left",
        row_names_side = "left",
        row_dend_side = "right",
        row_names_gp = grid::gpar(fontsize = 10),
        ## Column parameters
        cluster_columns = FALSE,
        column_title = "",
        column_title_gp = grid::gpar(fontsize = 12),
        column_names_side = "bottom",
        column_names_gp = grid::gpar(fontsize = 10),
        column_names_rot = 90,
        ## Legend parameters
        heatmap_legend_param = list(
          title = "Jaccard index",
          direction = "vertical",
          legend_width = unit(1.5, "in")
        ),
        show_heatmap_legend = celltype_method == "SingleR",
      )
    }
  ) |>
  # concatenate vertically into HeatmapList object
  purrr::reduce(ComplexHeatmap::`%v%`) |>
  ComplexHeatmap::draw(
    heatmap_legend_side = "right",
    # add a margin to the heatmap so labels don't get cut off
    padding = unit(c(2, 20, 2, 2), "mm")
  )
```

@@

```{r}
# prepare a data frame with one row per chromosome per cell for plotting
proportion_df <- cnv_df |>
  # select only relevant columns for plotting
  dplyr::select(barcodes, UMAP1, UMAP2, cnv_mean_classification, all_of(proportion_cols)) |>
  tidyr::pivot_longer(
    cols = all_of(proportion_cols),
    names_to = "chromosome",
    values_to = "proportion"
  ) |>
  # remove extra string in front of chr
  dplyr::mutate(
    chromosome = stringr::str_remove(chromosome, "proportion_scaled_cnv_chr"),
    # turn into numeric for easy ordering in plot
    chromosome = as.numeric(chromosome)
  )
```

```{r, fig.height=10, fig.width=10}
ggplot(proportion_df, aes(x = proportion, color = cnv_mean_classification)) +
  geom_density(bounds = c(0, 1), bw = 0.05) +
  theme_bw() +
  facet_wrap(vars(chromosome), ncol = 3)
```

@@

```{r}
ewings_chr_df <- proportion_df |>
  dplyr::filter(chromosome %in% c(1, 8, 12, 16))

ggplot(ewings_chr_df, aes(x = UMAP1, y = UMAP2, color = proportion)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw() +
  facet_wrap(vars(chromosome)) +
  scale_color_viridis_c()
```

### 2024-07-24 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd

```{r}
# create a combined classification between inferCNV and copyKAT
# cells that are classified as tumor by both are kept as tumor
# classified by both as normal are normal 
# in between are classified as ambiguous
classification_df <- classification_df |> 
  dplyr::mutate(
    combined_cnv_classification = dplyr::case_when(
      # only tumor if both call tumor 
      cnv_proportion_classification == "Tumor" & copykat_classification == "Tumor" ~ "Tumor",
      # only normal if both call normal and/or reference
      cnv_proportion_classification == "Normal" & copykat_classification == "Normal" ~ "Normal",
      cnv_proportion_classification == "Reference" & copykat_classification == "Normal" ~ "Normal",
      # if normal/tumor mismatch set to ambiguous 
      .default = "Ambiguous"
    ), 
    # set factor order to use for plotting
    combined_cnv_classification = forcats::fct_relevel(combined_cnv_classification, "Tumor", "Normal", "Ambiguous"))
```

@@

```{r}
# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- plot_markers_df |> 
  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
  tidyr::pivot_wider(values_from = gene_expression, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("gene_symbol") |> 
  as.matrix()

@@

annotation <- ComplexHeatmap::columnAnnotation(
  classification = annotation_df$combined_cnv_classification,
  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
```

```{r}
# plot heatmap of marker genes 
plot_gene_heatmap(marker_gene_heatmap, 
                  row_title = "Marker gene symbol",
                  legend_title = "Marker gene \nexpression",
                  annotation = annotation)
```

@@

```{r, fig.height=10}
geneset_plot_df |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset))
```

@@

```{r}
# make a matrix of gene set by barcode 
geneset_heatmap <- geneset_plot_df |> 
  dplyr::select(mean_score, geneset, barcodes) |> 
  unique() |> 
  tidyr::pivot_wider(values_from = mean_score, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("geneset") |> 
  as.matrix()

# plot heatmap of gene set score 
plot_gene_heatmap(geneset_heatmap,
                  annotation = annotation,
                  legend_title = "Gene set \nscore")
```

@@

```{r}
# need to make a new annotation, since we want to annotate the rows here 
annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_cnv_classification,
                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))

# plot gains in cnv
plot_cnv_heatmap(classification_df,
                 cnv_col = "proportion_scaled_dupli",
                 annotation = annotation,
                 legend_title = "Copy number \ngain")
```

@@

```{r}
# plot loss in cnv
plot_cnv_heatmap(classification_df, 
                 cnv_col = "proportion_scaled_loss", 
                 annotation = annotation, 
                 legend_title = "Copy number \nloss")
```

>>>>

```{r}
# create a combined classification between inferCNV and copyKAT
# cells that are classified as tumor by both are kept as tumor
# classified by both as normal are normal
# in between are classified as ambiguous
classification_df <- classification_df |>
  dplyr::mutate(
    combined_cnv_classification = dplyr::case_when(
      # only tumor if both call tumor
      cnv_proportion_classification == "Tumor" & copykat_classification == "Tumor" ~ "Tumor",
      # only normal if both call normal and/or reference
      cnv_proportion_classification == "Normal" & copykat_classification == "Normal" ~ "Normal",
      cnv_proportion_classification == "Reference" & copykat_classification == "Normal" ~ "Normal",
      # if normal/tumor mismatch set to ambiguous
      .default = "Ambiguous"
    ),
    # set factor order to use for plotting
    combined_cnv_classification = forcats::fct_relevel(combined_cnv_classification, "Tumor", "Normal", "Ambiguous")
  )
```

@@

```{r}
# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- plot_markers_df |>
  dplyr::select(gene_expression, gene_symbol, barcodes) |>
  tidyr::pivot_wider(
    values_from = gene_expression,
    names_from = barcodes
  ) |>
  tibble::column_to_rownames("gene_symbol") |>
  as.matrix()

@@

annotation <- ComplexHeatmap::columnAnnotation(
  classification = annotation_df$combined_cnv_classification,
  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"))
)
```

```{r}
# plot heatmap of marker genes
plot_gene_heatmap(marker_gene_heatmap,
  row_title = "Marker gene symbol",
  legend_title = "Marker gene \nexpression",
  annotation = annotation
)
```

@@

```{r, fig.height=10}
geneset_plot_df |>
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |>
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(
    rows = vars(method),
    cols = vars(geneset)
  )
```

@@

```{r}
# make a matrix of gene set by barcode
geneset_heatmap <- geneset_plot_df |>
  dplyr::select(mean_score, geneset, barcodes) |>
  unique() |>
  tidyr::pivot_wider(
    values_from = mean_score,
    names_from = barcodes
  ) |>
  tibble::column_to_rownames("geneset") |>
  as.matrix()

# plot heatmap of gene set score
plot_gene_heatmap(geneset_heatmap,
  annotation = annotation,
  legend_title = "Gene set \nscore"
)
```

@@

```{r}
# need to make a new annotation, since we want to annotate the rows here
annotation <- ComplexHeatmap::rowAnnotation(
  classification = annotation_df$combined_cnv_classification,
  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey"))
)

# plot gains in cnv
plot_cnv_heatmap(classification_df,
  cnv_col = "proportion_scaled_dupli",
  annotation = annotation,
  legend_title = "Copy number \ngain"
)
```

@@

```{r}
# plot loss in cnv
plot_cnv_heatmap(classification_df,
  cnv_col = "proportion_scaled_loss",
  annotation = annotation,
  legend_title = "Copy number \nloss"
)
```

### 2024-07-24 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd

```{r}
# generate classification df 
classification_df <- create_classification_df(
  sce,
  marker_gene_results_file,
  copykat_predictions_file, 
  infercnv_predictions_file,
  infercnv_metadata_file,
  geneset_scores_file
) |> 
  # remove extra infercnv classification column we won't use 
  dplyr::select(- cnv_sum_classification)
```

@@

```{r}
geneset_plot_df |> 
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |> 
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "copykat_classification" ~ "CopyKAT - no reference",
      method == "cnv_proportion_classification" ~ "InferCNV proportion"
    )) |> 
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(rows = vars(method), 
             cols = vars(geneset))
```

@@

# pull gene sets from msigbdr 
# all gene sets are part of C2, CGP
genes_df <- msigdbr::msigdbr(species= "Homo sapiens", 
                             category = "C2",
                             subcategory = "CGP") |> 
  # only keep relevant gene sets 
  dplyr::filter(gs_name %in% ews_gene_sets)
```

@@

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      caret_df$cnv_classification, 
      caret_df$auc_classification)
  ) 
```

@@

```{r}
ref_auc_df <- ref_auc@assays@data$AUC |> 
  as.data.frame() |> 
  tibble::rownames_to_column("gene_set") |> 
  tidyr::pivot_longer(!gene_set, 
                      names_to = "barcodes", 
                      values_to = "auc") |> 
  #dplyr::filter(gene_set == "markers") |> 
  dplyr::mutate(sample = "SCPCL000822")

new_auc_df <- new_auc@assays@data$AUC |> 
  as.data.frame() |> 
  tibble::rownames_to_column("gene_set") |> 
  tidyr::pivot_longer(!gene_set, 
                      names_to = "barcodes", 
                      values_to = "auc") |> 
  #dplyr::filter(gene_set == "markers") |> 
  dplyr::mutate(sample = "SCPCL000824")

@@

ggplot(all_auc_df, aes(x = auc, color = sample)) +
  geom_density() +
  facet_grid(rows = vars(gene_set),
             scales = "free_y")
```

@@

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      classification_df$updated_marker_gene_classification, 
      classification_df$auc_classification)
  ) 
```

@@

ucell_df <- new_ucell |> 
  as.data.frame() |> 
  tibble::rownames_to_column("barcodes") |> 
  dplyr::mutate(sample = "SCPCL000824") |> 
  dplyr::bind_rows(ref_ucell) |> 
  tidyr::pivot_longer(ends_with("UCell"), 
                      names_to = "gene_list", 
                      values_to = "signature_score") |> 
  dplyr::mutate(
    gene_list = stringr::word(gene_list, 1, sep = "_")
  )
```

```{r}
ggplot(ucell_df, aes(x = signature_score, colour = sample)) +
  geom_density() +
  facet_grid(rows = vars(gene_list),
             scales = "free_y")
```

@@

# get a vector of tumor cells from ref 
ref_ucell_tumor_cells <- ucell_df |> 
  dplyr::filter(gene_list == "markers",
                signature_score >= geneset_cutoff,
                sample == "SCPCL000822") |> 
  dplyr::pull(barcodes) |> 
  unique()

@@

```{r}
# get list of tumor cells using ucell 
new_ucell_tumor_cells <- ucell_df |> 
  dplyr::filter(gene_list == "markers",
                signature_score >= geneset_cutoff,
                sample == "SCPCL000824") |> 
  dplyr::pull(barcodes) |> 
  unique()

@@

# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- plot_markers_df |> 
  dplyr::select(gene_expression, gene_symbol, barcodes) |> 
  tidyr::pivot_wider(values_from = gene_expression, 
                     names_from = barcodes) |> 
  tibble::column_to_rownames("gene_symbol") |> 
  as.matrix()

@@

annotation <- ComplexHeatmap::columnAnnotation(
  marker_genes = annotation_df$marker_gene_classification,
  updated_marker_genes = annotation_df$updated_marker_gene_classification,
  AUCell = annotation_df$auc_classification, 
  UCell = annotation_df$ucell_classification,
  consensus = annotation_df$consensus_classification,
  col = list(marker_genes = plot_colors,
             updated_marker_genes =plot_colors, 
             AUCell = plot_colors,
             UCell = plot_colors,
             consensus = plot_colors))
```

```{r}
# plot heatmap of marker genes 
plot_gene_heatmap(marker_gene_heatmap, 
                  row_title = "Marker gene symbol",
                  legend_title = "Marker gene \nexpression",
                  annotation = annotation)
```

>>>>

```{r}
# generate classification df
classification_df <- create_classification_df(
  sce,
  marker_gene_results_file,
  copykat_predictions_file,
  infercnv_predictions_file,
  infercnv_metadata_file,
  geneset_scores_file
) |>
  # remove extra infercnv classification column we won't use
  dplyr::select(-cnv_sum_classification)
```

@@

```{r}
geneset_plot_df |>
  tidyr::pivot_longer(
    cols = ends_with("classification"),
    names_to = "method",
    values_to = "classification"
  ) |>
  dplyr::mutate(
    method = dplyr::case_when(
      method == "marker_gene_classification" ~ "Marker genes only",
      method == "copykat_classification" ~ "CopyKAT - no reference",
      method == "cnv_proportion_classification" ~ "InferCNV proportion"
    )
  ) |>
  ggplot(aes(x = mean_score, color = classification)) +
  geom_density(bw = 0.05) +
  facet_grid(
    rows = vars(method),
    cols = vars(geneset)
  )
```

@@

# pull gene sets from msigbdr
# all gene sets are part of C2, CGP
genes_df <- msigdbr::msigdbr(
  species = "Homo sapiens",
  category = "C2",
  subcategory = "CGP"
) |>
  # only keep relevant gene sets
  dplyr::filter(gs_name %in% ews_gene_sets)
```

@@

# compare using a confusion matrix
caret::confusionMatrix(
  table(
    caret_df$cnv_classification,
    caret_df$auc_classification
  )
)
```

@@

```{r}
ref_auc_df <- ref_auc@assays@data$AUC |>
  as.data.frame() |>
  tibble::rownames_to_column("gene_set") |>
  tidyr::pivot_longer(!gene_set,
    names_to = "barcodes",
    values_to = "auc"
  ) |>
  # dplyr::filter(gene_set == "markers") |>
  dplyr::mutate(sample = "SCPCL000822")

new_auc_df <- new_auc@assays@data$AUC |>
  as.data.frame() |>
  tibble::rownames_to_column("gene_set") |>
  tidyr::pivot_longer(!gene_set,
    names_to = "barcodes",
    values_to = "auc"
  ) |>
  # dplyr::filter(gene_set == "markers") |>
  dplyr::mutate(sample = "SCPCL000824")

@@

ggplot(all_auc_df, aes(x = auc, color = sample)) +
  geom_density() +
  facet_grid(
    rows = vars(gene_set),
    scales = "free_y"
  )
```

@@

# compare using a confusion matrix
caret::confusionMatrix(
  table(
    classification_df$updated_marker_gene_classification,
    classification_df$auc_classification
  )
)
```

@@

ucell_df <- new_ucell |>
  as.data.frame() |>
  tibble::rownames_to_column("barcodes") |>
  dplyr::mutate(sample = "SCPCL000824") |>
  dplyr::bind_rows(ref_ucell) |>
  tidyr::pivot_longer(ends_with("UCell"),
    names_to = "gene_list",
    values_to = "signature_score"
  ) |>
  dplyr::mutate(
    gene_list = stringr::word(gene_list, 1, sep = "_")
  )
```

```{r}
ggplot(ucell_df, aes(x = signature_score, colour = sample)) +
  geom_density() +
  facet_grid(
    rows = vars(gene_list),
    scales = "free_y"
  )
```

@@

# get a vector of tumor cells from ref
ref_ucell_tumor_cells <- ucell_df |>
  dplyr::filter(
    gene_list == "markers",
    signature_score >= geneset_cutoff,
    sample == "SCPCL000822"
  ) |>
  dplyr::pull(barcodes) |>
  unique()

@@

```{r}
# get list of tumor cells using ucell
new_ucell_tumor_cells <- ucell_df |>
  dplyr::filter(
    gene_list == "markers",
    signature_score >= geneset_cutoff,
    sample == "SCPCL000824"
  ) |>
  dplyr::pull(barcodes) |>
  unique()

@@

# create matrix with marker genes as rows and barcodes as columns
marker_gene_heatmap <- plot_markers_df |>
  dplyr::select(gene_expression, gene_symbol, barcodes) |>
  tidyr::pivot_wider(
    values_from = gene_expression,
    names_from = barcodes
  ) |>
  tibble::column_to_rownames("gene_symbol") |>
  as.matrix()

@@

annotation <- ComplexHeatmap::columnAnnotation(
  marker_genes = annotation_df$marker_gene_classification,
  updated_marker_genes = annotation_df$updated_marker_gene_classification,
  AUCell = annotation_df$auc_classification,
  UCell = annotation_df$ucell_classification,
  consensus = annotation_df$consensus_classification,
  col = list(
    marker_genes = plot_colors,
    updated_marker_genes = plot_colors,
    AUCell = plot_colors,
    UCell = plot_colors,
    consensus = plot_colors
  )
)
```

```{r}
# plot heatmap of marker genes
plot_gene_heatmap(marker_gene_heatmap,
  row_title = "Marker gene symbol",
  legend_title = "Marker gene \nexpression",
  annotation = annotation
)
```

### 2024-07-24 analyses/cell-type-ewings/template_notebooks/cnv-workflow/03-copykat.Rmd
fix up some of the ewings files with diffs

# check if marker gene annotations are present 
if(all(is.na(manual_classifications_df$marker_gene_classification))){
  has_marker_gene <- FALSE
  message("No annotations were available using only marker gene expression. 
          Any plots comparing CopyKAT to marker gene annotation will be skipped.")
} else {
  has_marker_gene <- TRUE 
}

@@

cnv_df <- umap_df |> 
  # first add manual annotations
  dplyr::left_join(manual_classifications_df) |> 
  # now add copykat results
  dplyr::left_join(ck_results_df, by = c("barcodes" = "cell.names")) |> 
  dplyr::mutate(
    copykat.pred = dplyr::if_else(
      # label any reference cells that were used 
      (barcodes %in% normal_cells) & (reference_used == "with_ref"), 
      "reference", 
      copykat.pred)
  )
```

@@

```{r, eval=has_marker_gene}
filtered_cnv_df <- cnv_df |> 
  dplyr::filter(!(copykat.pred %in% c("not.defined", "reference")),
                # filter any low confidence calls
                !stringr::str_detect(copykat.pred, "low.conf"))

@@

```{r, eval=has_marker_gene}
# calculate Jaccard similarity index for each reference type
jaccard_matrices <- caret_df_list |>
  purrr::map(\(df) {

    make_jaccard_matrix(
      df,
      "marker_gene_classification",
      "copykat.pred"
    )
  })
```

@@

```{r}
celltype_df <- cnv_df |> 
  dplyr::select(barcodes, 
                reference_used,
                copykat.pred) |> 
  dplyr::distinct() |> 
  tidyr::pivot_wider(
    names_from = reference_used,
    values_from = copykat.pred
  )

all_info_df <- cnv_df |> 
  dplyr::select(barcodes,
                mean_cnv_detection) |> 
  dplyr::group_by(barcodes) |>
  dplyr::mutate(mean_cnv_detection = mean(mean_cnv_detection)) |> 
  dplyr::distinct() |> 
  dplyr::left_join(celltype_df)

>>>>

# check if marker gene annotations are present
if (all(is.na(manual_classifications_df$marker_gene_classification))) {
  has_marker_gene <- FALSE
  message("No annotations were available using only marker gene expression.
          Any plots comparing CopyKAT to marker gene annotation will be skipped.")
} else {
  has_marker_gene <- TRUE
}

@@

cnv_df <- umap_df |>
  # first add manual annotations
  dplyr::left_join(manual_classifications_df) |>
  # now add copykat results
  dplyr::left_join(ck_results_df, by = c("barcodes" = "cell.names")) |>
  dplyr::mutate(
    copykat.pred = dplyr::if_else(
      # label any reference cells that were used
      (barcodes %in% normal_cells) & (reference_used == "with_ref"),
      "reference",
      copykat.pred
    )
  )
```

@@

```{r, eval=has_marker_gene}
filtered_cnv_df <- cnv_df |>
  dplyr::filter(
    !(copykat.pred %in% c("not.defined", "reference")),
    # filter any low confidence calls
    !stringr::str_detect(copykat.pred, "low.conf")
  )

@@

```{r, eval=has_marker_gene}
# calculate Jaccard similarity index for each reference type
jaccard_matrices <- caret_df_list |>
  purrr::map(\(df) {
    make_jaccard_matrix(
      df,
      "marker_gene_classification",
      "copykat.pred"
    )
  })
```

@@

```{r}
celltype_df <- cnv_df |>
  dplyr::select(
    barcodes,
    reference_used,
    copykat.pred
  ) |>
  dplyr::distinct() |>
  tidyr::pivot_wider(
    names_from = reference_used,
    values_from = copykat.pred
  )

all_info_df <- cnv_df |>
  dplyr::select(
    barcodes,
    mean_cnv_detection
  ) |>
  dplyr::group_by(barcodes) |>
  dplyr::mutate(mean_cnv_detection = mean(mean_cnv_detection)) |>
  dplyr::distinct() |>
  dplyr::left_join(celltype_df)
